
# Analysis of the Data from INEGI
INEGI is the Mexican National Institute of Statistics and Geography, which provides a wealth of data on various aspects of Mexican society, including demographics, economics, and geography. The data we will be analyzing comes from the 2020 and 2010 census datasets for Yucatán.

We are particularly interested in how migration patterns have affected the population of the biggest cities in the state, Merida, Valladolid and the coast area, like Progreso and Telchac.


```{r}
file_path_2020 <- "/Users/ecastillo/Documents/coding/analyzing_merida_data/ITER2020 - 31 Yucatán.csv"
data_2020 <- read.csv(file_path_2020, fileEncoding = "UTF-8", sep = ",", header = TRUE)
file_path_2010 <- "/Users/ecastillo/Documents/coding/analyzing_merida_data/RESLOC2010 - 31 Yucatán.csv" 
data_2010 <- read.csv(file_path_2010, fileEncoding = "UTF-8", sep = ",", header = TRUE)
# Display the first few rows of the dataset
head(data_2020, 10)
# Display the first few rows of the dataset
head(data_2010, 10)
```

We also need a mapping between the column names and its meaning:
```{r}
# Create a mapping of column names to their meanings
query_to_readable_en_2020 <- c(
  "ENTIDAD" = "State code",
  "NOM_ENT" = "State name",
  "MUN" = "Municipality or borough code",
  "NOM_MUN" = "Municipality or borough name",
  "LOC" = "Locality code",
  "NOM_LOC" = "Locality name",
  "POBTOT" = "Total population",
  "PNACENT" = "Population born in the state",
  "PNACOE" = "Population born in another state",
  "P3YM_HLI" = "Population aged 3+ speaking an Indigenous language",
  "P3HLINHE" = "Population aged 3+ speaking an Indigenous language and not Spanish",
  "P3HLI_HE" = "Population aged 3+ speaking an Indigenous language and Spanish",
  "P5_HLI" = "Population aged 5+ speaking an Indigenous language",
  "P5_HLI_NHE" = "Population aged 5+ speaking an Indigenous language and not Spanish",
  "P5_HLI_HE" = "Population aged 5+ speaking an Indigenous language and Spanish",
  "PHOG_IND" = "Population in Indigenous census households",
  "POB_AFRO" = "Population identifying as Afro-Mexican or of African descent",
  "GRAPROES" = "Average years of schooling",
  "PEA" = "Economically active population aged 12+",
  "POCUPADA" = "Employed population aged 12+",
  "PDESOCUP" = "Unemployed population aged 12+",
  "PE_INAC" = "Economically inactive population aged 12+",
  "TOTHOG" = "Total census households",
  "POBHOG" = "Population in census households",
  "VIVTOT" = "Total housing units",
  "TVIVHAB" = "Total inhabited housing units",
  "TVIVPAR" = "Total private housing units",
  "VIVPAR_DES" = "Uninhabited private housing units",
  "VIVPAR_UT" = "Private housing units for seasonal use",
  "OCUPVIVPAR" = "Occupants in inhabited private housing units",
  "TVIVPARHAB" = "Total inhabited private housing units",
  "VIVPAR_HAB" = "Inhabited private housing units",
  "PROM_OCUP" = "Average occupants per inhabited private housing unit",
  "PRO_OCUP_C" = "Average occupants per room in inhabited private housing units",
  "PSINDER" = "Population without health service affiliation",
  "PDER_SS" = "Population with health service affiliation"
)

query_to_readable_en_2010 <- c(
  "ENTIDAD"       = "State code",
  "NOM_ENT"       = "State name",
  "MUN"           = "Municipality or borough code",
  "NOM_MUN"       = "Municipality or borough name",
  "LOC"           = "Locality code",
  "NOM_LOC"       = "Locality name",
  "P_TOTAL"       = "Total population",
  "TOTHOG"        = "Total census households",
  "P3YM_HLI"      = "Population aged 3+ speaking an Indigenous language",
  "P3HLINHE"      = "Population aged 3+ speaking an Indigenous language and not Spanish",
  "P3HLI_HE"      = "Population aged 3+ speaking an Indigenous language and Spanish",
  "P5_HLI"        = "Population aged 5+ speaking an Indigenous language",
  "P5_HLI_NHE"    = "Population aged 5+ speaking an Indigenous language and not Spanish",
  "P5_HLI_HE"     = "Population aged 5+ speaking an Indigenous language and Spanish",
  "PHOG_IND"      = "Population in Indigenous census households",
  "PNACENT"       = "Population born in the state",
  "PNACOE"        = "Population born in another state",
  "VIVTOT"        = "Total housing units",
  "T_VIVHAB"      = "Total inhabited housing units",
  "TVIVPAR"       = "Total private housing units",
  "VIVPAR_HAB"    = "Inhabited private housing units",
  "TVIVPARHAB"    = "Total inhabited private housing units"
)
```

# Analysis of the Data from INEGI
We will analyze the data to understand the migration patterns and demographic changes in Yucatán, focusing on the cities of Merida, Valladolid, Progreso, and Telchac. The analysis will include:
- Population changes between 2010 and 2020.
- How has the migration impacted the density of Indigenous people in these cities.
- How has this impacted the occupation of houses.

## Lets explore the data
```{r}
# Display the structure of the 2020 dataset
str(data_2020)
# Display the structure of the 2010 dataset
str(data_2010)
```

We can see that the data types may not be correct, so we will convert them when needed.

## Let's see the distribution of the population in the state of Yucatán
```{r}
library(ggplot2)
library(dplyr)
# lets see which are the biggest cities in Yucatán, the data has the total population as MUN == 0, so lets filter that out
# the LOC = 0 is the total of the municipality
# we only want to keep the
biggest_cities_2020 <- data_2020 %>%
  filter(MUN != 0) %>%
  filter(LOC == 0) %>%
  arrange(desc(POBTOT))

# Lets add the population from 2010
biggest_cities_2010 <- data_2010 %>%
  filter(MUN != 0) %>%
  filter(LOC == 0) %>%
  arrange(desc(P_TOTAL))
```

Now that we have only the totals for each municipality, lets make sure we have the 106 Municipalities that exist in Yucatán.
```{r}
# Check the number of unique municipalities in the 2020 data
num_municipalities_2020 <- nrow(biggest_cities_2020)
num_municipalities_2010 <- nrow(biggest_cities_2010)
cat("Number of unique municipalities in 2020:", num_municipalities_2020, "\n")
cat("Number of unique municipalities in 2010:", num_municipalities_2010, "\n")
```

We can see that there are 106 municipalities in both datasets, so we can proceed with the analysis.

## What are the biggest cities in Yucatan at the moment (2020)?
```{r}
# the total population is in the row where MUN is 0 and LOC is 0, so we can filter that out
total_population_2020 <- (
  data_2020 %>%
  filter(MUN == 0, LOC == 0) %>%
  select(POBTOT)
  )$POBTOT



# Display the top 10 biggest cities in Yucatán in 2020, add a column with the percentage of the total population
biggest_cities_2020 %>%
  select(NOM_MUN, POBTOT) %>%
  arrange(desc(POBTOT)) %>%
  mutate(PercentageOfTotal = POBTOT / total_population_2020 * 100) %>%
  head(10)
```

## Lets see the distribution of the population in the biggest cities
Lets analyze the population that was born in the state vs in another state with a pie graph.
```{r}
pop_born_in_state <- biggest_cities_2020 %>%
  select(NOM_MUN, POBTOT, PNACENT, PNACOE) %>%
  mutate(
    BornInState = PNACENT,
    BornInOtherState = PNACOE
  ) %>%
  select(NOM_MUN, POBTOT, BornInState, BornInOtherState) %>%
  arrange(desc(POBTOT))
```
## What were the biggest cities in Yucatan in 2010?


## Population Changes
Let's see which cities had the biggest changes in population between 2010 and 2020.
For that we need to focus only on total population, population born in the state and population born in another state.
```{r}
population_changes <- biggest_cities_2020 %>%
  select(MUN, NOM_MUN, POBTOT, PNACENT, PNACOE) %>%
  rename(
    Population2020 = POBTOT,
    BornInState2020 = PNACENT,
    BornInOtherState2020 = PNACOE
  ) %>%
  # convert to numeric
  mutate(
    Population2020 = as.numeric(Population2020),
    BornInState2020 = as.numeric(BornInState2020),
    BornInOtherState2020 = as.numeric(BornInOtherState2020)
  ) %>%
  left_join(
    biggest_cities_2010 %>%
      select(MUN, P_TOTAL, PNACENT, PNACOE) %>%
      rename(Population2010 = P_TOTAL,
             BornInState2010 = PNACENT,
             BornInOtherState2010 = PNACOE) %>%
      mutate(
        MUN = as.numeric(MUN),
        Population2010 = as.numeric(Population2010),
        BornInState2010 = as.numeric(BornInState2010),
        BornInOtherState2010 = as.numeric(BornInOtherState2010)
      ),
    by = "MUN"
  ) %>%
  mutate(
    MUN = as.numeric(MUN),
    PopulationChange = Population2020 - Population2010,
    BornInStateChange = BornInState2020 - BornInState2010,
    BornInOtherStateChange = BornInOtherState2020 - BornInOtherState2010,
    PercentageChange = (PopulationChange / Population2010) * 100
  ) %>%
  arrange(desc(PopulationChange))

```
## Display the population changes

Total population changes.
```{r}
library(knitr)
kable(population_changes %>%
  select(MUN, NOM_MUN, Population2010, Population2020, PopulationChange, PercentageChange,) %>%
  arrange(desc(PercentageChange)),
  caption = "Population Changes in Yucatán Municipalities (2010–2020)",
  digits = 2)
```

Plot it in a bar graph, only the top 10 biggest changes, do it in percentage with respect to 2010. Increase the resolution of the plot.
```{r}
library(ggplot2)
plot <- ggplot(population_changes %>%
         select(MUN, NOM_MUN, Population2010, Population2020, PopulationChange, PercentageChange) %>%
         arrange(desc(PercentageChange)) %>%
         head(10),
       aes(x = reorder(NOM_MUN, PercentageChange), y = PercentageChange)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Population Changes in Yucatán Municipalities (2010–2020)",
    x = "Municipality",
    y = "Percentage Change (%)"
  ) +
  theme_minimal()
ggsave("population_changes_top_10.png", plot, width = 10, height = 6, dpi = 300)
print(plot)
```

Born in state changes. Save the top ten of the table to a file.
```{r}
total_population_2010 <- (
  data_2010 %>%
  filter(MUN == 0, LOC == 0) %>%
  select(P_TOTAL)
  )$P_TOTAL

table <- kable(population_changes %>%
  mutate(
    PercentageChange = round((BornInStateChange / BornInState2010) * 100, 2),
    PercentageChangeToMunicipalityPop = round((BornInStateChange / Population2010) * 100, 2),
    PercentageChangeToTotalPop = round((BornInStateChange / total_population_2010) * 100, 2),
  ) %>%
  select(MUN, NOM_MUN, BornInState2010, BornInState2020, BornInStateChange, PercentageChange, PercentageChangeToMunicipalityPop, PercentageChangeToTotalPop) %>%
  arrange(desc(PercentageChangeToMunicipalityPop))-> population_changes_in_state,
  caption = "Population Changes Born in Yucatán (2010–2020)",
  digits = 2)
# Save the table to a file
write.csv(population_changes_in_state %>%
  select(MUN, NOM_MUN, BornInState2010, BornInState2020, BornInStateChange, PercentageChange, PercentageChangeToMunicipalityPop, PercentageChangeToTotalPop) %>%
  arrange(desc(PercentageChangeToMunicipalityPop)),
  "population_changes_born_in_state.csv", row.names = FALSE, )
print(table)
```
Born in other state changes.
```{r}
table <- kable(population_changes %>%
  mutate(
    PercentageChange = round(((BornInOtherStateChange / BornInOtherState2010) * 100), 2),
    PercentageChangeToMunicipalityPop = round((BornInOtherStateChange / Population2010) * 100, 2),
    PercentageChangeToTotalPop = round((BornInOtherStateChange / total_population_2010) * 100, 2),        
  ) %>%
  select(MUN, NOM_MUN, BornInOtherState2010, BornInOtherState2020, BornInOtherStateChange, PercentageChange, PercentageChangeToMunicipalityPop, PercentageChangeToTotalPop) %>%
  arrange(desc(PercentageChangeToMunicipalityPop)) -> population_changes_out_state,
  caption = "Population Changes Born in Other States (2010–2020)",
  digits = 2)

  # Save the table to a file
write.csv(population_changes_out_state %>%
  select(MUN, NOM_MUN, BornInOtherState2010, BornInOtherState2020, BornInOtherStateChange, PercentageChange, PercentageChangeToMunicipalityPop, PercentageChangeToTotalPop) %>%
  arrange(desc(PercentageChangeToMunicipalityPop)),
  "population_changes_born_in_other_state.csv", row.names = FALSE)
print(table)
```



## Let's visualize the changes in population 
Lets create a plot with the changes in population on top of a map.
```{r}
library(sf)
yucatan_map <- st_read("/Users/ecastillo/Documents/coding/analyzing_merida_data/yucatan.geojson")
yucatan_map <- yucatan_map %>%
  mutate(CVE_MUN = as.numeric(CVE_MUN)) # Ensure CVE_MUN is numeric for joining
map_data <- yucatan_map %>%
  left_join(population_changes, by = c("CVE_MUN" = "MUN"))

# Calculate centroids
map_data_centroids <- st_centroid(map_data)
map_plot = ggplot(data = map_data) +
  geom_sf(aes(fill = PopulationChange), color = "white") +
  geom_sf_text(data = map_data_centroids, aes(label = NOMGEO), size = 2.5, color = "black") +
  scale_fill_viridis_c(option = "C") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Population Change by Municipality (2010–2020)",
    fill = "Pop. Change"
  )
print(map_plot)
# save to png
ggsave("population_change_map.png", map_plot, width = 10, height = 6, dpi = 300)
```

We can see that the biggest increases in population are around the areas of Merida, Valladolid and Tizimin.

Lets see if we can find the same pattern in the influx of people from outside the state.

## Let's visualize the changes in population born in Yucatan state and in other states.

```{r}
# we need to get the max and min values for the fill scale
global_min <- min(map_data$BornInStateChange, map_data$BornInOtherStateChange, na.rm = TRUE)
global_max <- max(map_data$BornInStateChange, map_data$BornInOtherStateChange, na.rm = TRUE)

```
```{r}
map_plot = ggplot(data = map_data) +
  geom_sf(aes(fill = BornInStateChange), color = "white") +
  geom_sf_text(data = map_data_centroids, aes(label = NOMGEO), size = 2.5, color = "black") +
  scale_fill_viridis_c(option = "C", limits = c(global_min, global_max)) +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Population Change from Within Yucatán (2010–2020)",
    fill = "Pop. Change"
  )
print(map_plot)
# save to png
ggsave("population_change_born_in_state_map.png", map_plot, width = 10, height = 6, dpi = 300)
```

```{r}

map_plot = ggplot(data = map_data) +
  geom_sf(aes(fill = BornInOtherStateChange), color = "white") +
  geom_sf_text(data = map_data_centroids, aes(label = NOMGEO), size = 2.5, color = "black") +
  scale_fill_viridis_c(option = "C", limits = c(global_min, global_max)) +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Population Change from Other States (2010–2020)",
    fill = "Pop. Change"
  )
print(map_plot)
# save to png
ggsave("population_change_born_in_other_state_map.png", map_plot, width = 10, height = 6, dpi = 300)
```

The trend is similar.

## Let's see how the Indigenous population has changed
```{r}
indigenous_population_changes <- biggest_cities_2020 %>%
  select(MUN, NOM_MUN, P3YM_HLI, P3HLINHE, P3HLI_HE, P5_HLI, P5_HLI_NHE, P5_HLI_HE) %>%
  rename(
    IndigenousPopulation2020 = P3YM_HLI,
    IndigenousNotSpanish2020 = P3HLINHE,
    IndigenousAndSpanish2020 = P3HLI_HE,
    IndigenousPopulation5Plus2020 = P5_HLI,
    IndigenousNotSpanish5Plus2020 = P5_HLI_NHE,
    IndigenousAndSpanish5Plus2020 = P5_HLI_HE
  ) %>%
  mutate(
    MUN = as.numeric(MUN),
    IndigenousPopulation2020 = as.numeric(IndigenousPopulation2020),
    IndigenousNotSpanish2020 = as.numeric(IndigenousNotSpanish2020),
    IndigenousAndSpanish2020 = as.numeric(IndigenousAndSpanish2020),
    IndigenousPopulation5Plus2020 = as.numeric(IndigenousPopulation5Plus2020),
    IndigenousNotSpanish5Plus2020 = as.numeric(IndigenousNotSpanish5Plus2020),
    IndigenousAndSpanish5Plus2020 = as.numeric(IndigenousAndSpanish5Plus2020)
  ) %>%
  left_join(
    biggest_cities_2010 %>%
      select(MUN, P3YM_HLI, P3HLINHE, P3HLI_HE, P5_HLI, P5_HLI_NHE, P5_HLI_HE) %>%
      rename(
        IndigenousPopulation2010 = P3YM_HLI,
        IndigenousNotSpanish2010 = P3HLINHE,
        IndigenousAndSpanish2010 = P3HLI_HE,
        IndigenousPopulation5Plus2010 = P5_HLI,
        IndigenousNotSpanish5Plus2010 = P5_HLI_NHE,
        IndigenousAndSpanish5Plus2010 = P5_HLI_HE
      ) %>%
      mutate(
        MUN = as.numeric(MUN),
        IndigenousPopulation2010 = as.numeric(IndigenousPopulation2010),
        IndigenousNotSpanish2010 = as.numeric(IndigenousNotSpanish2010),
        IndigenousAndSpanish2010 = as.numeric(IndigenousAndSpanish2010),
        IndigenousPopulation5Plus2010 = as.numeric(IndigenousPopulation5Plus2010),
        IndigenousNotSpanish5Plus2010 = as.numeric(IndigenousNotSpanish5Plus2010),
        IndigenousAndSpanish5Plus2010 = as.numeric(IndigenousAndSpanish5Plus2010)
      ),
    by = "MUN"
  ) %>%
  mutate(
    MUN = as.numeric(MUN),
    IndigenousPopulationChange = IndigenousPopulation2020 - IndigenousPopulation2010,
    IndigenousNotSpanishChange = IndigenousNotSpanish2020 - IndigenousNotSpanish2010,
    IndigenousAndSpanishChange = IndigenousAndSpanish2020 - IndigenousAndSpanish2010,
    IndigenousPopulation5PlusChange = IndigenousPopulation5Plus2020 - IndigenousPopulation5Plus2010,
    IndigenousNotSpanish5PlusChange = IndigenousNotSpanish5Plus2020 - IndigenousNotSpanish5Plus2010,
    IndigenousAndSpanish5PlusChange = IndigenousAndSpanish5Plus2020 - IndigenousAndSpanish5Plus2010
  ) %>%
  arrange(desc(IndigenousPopulationChange))
# Display the Indigenous population changes
kable(indigenous_population_changes %>%
  select(MUN, NOM_MUN, IndigenousPopulation2010, IndigenousPopulation2020, IndigenousPopulationChange),
  caption = "Indigenous Population Changes in Yucatán Municipalities (2010–2020)",
  digits = 2)
```

## Let's visualize the changes in Indigenous population
```{r}

map_data <- yucatan_map %>%
  left_join(indigenous_population_changes, by = c("CVE_MUN" = "MUN"))
map_plot <- ggplot(data = map_data) +
  geom_sf(aes(fill = IndigenousPopulationChange), color = "white") +
  geom_sf_text(data = map_data_centroids, aes(label = NOMGEO), size = 2.5, color = "black") +
  scale_fill_viridis_c(option = "C") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Indigenous Population Change by Municipality (2010–2020)",
    fill = "Pop. Change"
  )
print(map_plot)
# save to png
ggsave("indigenous_population_change_map.png", map_plot, width = 10, height = 6, dpi = 300)
```

We can see that there has been a decrease in the Indigenous population in Merida. A municipality that saw a significant increase is Kanasin, which sits next to Merida. 
Kanasin is singificantly more affordable than Merida, so it is likely that indigenous people are moving there.


```{r}
ggplot(population_changes %>%
         left_join(indigenous_population_changes, by = "MUN"),
       aes(x = BornInOtherStateChange, y = IndigenousPopulationChange)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Relationship Between Out-of-State Migration and Indigenous Population Change",
    x = "Change in Population Born in Other States",
    y = "Change in Indigenous Population"
  ) +
  theme_minimal()
```
This plot shows the relationship between the change in population born in other states and the change in Indigenous population. The trend line suggests that as more people from other states move to Yucatán, there is a corresponding decrease in the Indigenous population.
But it has one outlier that is skewing the results, so we will first see what it is and then we will remove it and replot it.

```{r}
# Identify the outlier
# join, but keep only one NOM_MUN column
population_changes %>%
  left_join(indigenous_population_changes, by = "MUN") %>%
  mutate(
    ratio = BornInOtherStateChange / IndigenousPopulationChange
  ) %>%
  arrange(desc(ratio)) %>%
  rename(
    NOM_MUN = NOM_MUN.x
  ) %>%
  select(MUN, NOM_MUN, BornInOtherStateChange, IndigenousPopulationChange, ratio) %>%
  tail(10)
```

Conkal is the outlier, it has a very high increase in population born in other states, but a very low increase in Indigenous population. This is likely due to the fact that Conkal is one of the most popular places to move into Merida, it is just next 
to the city of Merida, a lot of new housing developments both luxurious and affordable.

Merida is also skewing the results. Kanasin as well.

Let's remove it to see the plot again more clearly.
Add names to the points.

```{r}
population_changes %>%
  left_join(indigenous_population_changes, by = "MUN") %>%
  filter(MUN != 13, MUN != 41, MUN != 50, MUN !=59, MUN != 101, MUN != 102, MUN != 96) %>% # remove Merida, Kanasin, Conkal
  # remove Conkal, Merida, Dzilam de Bravo
  ggplot(aes(x = BornInOtherStateChange, y = IndigenousPopulationChange)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  geom_text(aes(label = NOM_MUN.x), vjust = -0.5, size = 3) +
  labs(
    title = "Relationship Between Out-of-State Migration and Indigenous Population Change (Without major cities)",
    x = "Change in Population Born in Other States",
    y = "Change in Indigenous Population"
  ) +
  theme_minimal()
```

The plot with the removed entries. 

```{r}
population_changes %>%
  left_join(indigenous_population_changes, by = "MUN") %>%
  filter(MUN == 13 | MUN == 41 | MUN == 50 | MUN ==59 | MUN == 101 | MUN == 102 | MUN == 96) %>% 
  ggplot(aes(x = BornInOtherStateChange, y = IndigenousPopulationChange)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  geom_text(aes(label = NOM_MUN.x), vjust = -0.5, size = 3) +
  labs(
    title = "Relationship Between Out-of-State Migration and Indigenous Population Change (Only Major cities)",
    x = "Change in Population Born in Other States",
    y = "Change in Indigenous Population"
  ) +
  theme_minimal()
```

```{r}
map_data <- yucatan_map %>%
  mutate(CVE_MUN = as.numeric(CVE_MUN)) %>%
  left_join(indigenous_population_changes, by = c("CVE_MUN" = "MUN")) %>%
  left_join(population_changes, by = c("CVE_MUN" = "MUN")) %>%
  mutate(
    IndigenousShare2010 = IndigenousPopulation2010 / Population2010 * 100,
    IndigenousShare2020 = IndigenousPopulation2020 / Population2020 * 100
  )

# get the min and max values for the fill scale
global_min <- min(map_data$IndigenousShare2010, map_data$IndigenousShare2020, na.rm = TRUE)
global_max <- max(map_data$IndigenousShare2010, map_data$IndigenousShare2020, na.rm = TRUE)

# Now it has geometry + your data, so this will work
map_plot <- ggplot(map_data) +
  geom_sf(aes(fill = IndigenousShare2010), color = "white") +
  scale_fill_viridis_c(option = "D", limits = c(global_min, global_max)) +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 22.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "% of Indigenous Population in 2010",
    fill = "% Indigenous"
  )
print(map_plot)
ggsave("indigenous_population_share_2010.png", map_plot, width = 10, height = 6, dpi = 300)
```



```{r}

# Now it has geometry + your data, so this will work
map_plot <- ggplot(map_data) +
  geom_sf(aes(fill = IndigenousShare2020), color = "white") +
  scale_fill_viridis_c(option = "D", limits = c(global_min, global_max)) +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 22.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "% of Indigenous Population in 2020",
    fill = "% Indigenous"
  )
print(map_plot)
ggsave("indigenous_population_share_2020.png", map_plot, width = 10, height = 6, dpi = 300)
```

This shows a trend, the Indigenous population is decreasing in the state of Yucatán. either by being displaced or because their roots are being lost, the metholodgy of INEGI allows people to identify themselves.
Another thing that is clear to see is that the indigenous population is concentrated in the south of the state, and from 2010 to 2020 the trend becomer more clear.

Just out of curiosity, lets see how the change in Indigenous population that speak a native language has changed.
```{r}

map_plot <- ggplot(map_data) +
  geom_sf(aes(fill = IndigenousPopulation5PlusChange), color = "white") +
  geom_sf_text(data = map_data_centroids, aes(label = NOMGEO), size = 2.5, color = "black") +
  scale_fill_viridis_c(option = "C") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Native language speaking Population Change by Municipality (2010–2020)",
    fill = "Pop. Change"
  )
print(map_plot)
# save to png
ggsave("native_language_population_change_map.png", map_plot, width = 10, height = 6, dpi = 300)
```

# Let´s do the same with out of state migration
```{r}
# Start from yucatan_map to preserve geometry
map_data <- yucatan_map %>%
  mutate(CVE_MUN = as.numeric(CVE_MUN)) %>%
  left_join(population_changes, by = c("CVE_MUN" = "MUN")) %>%
  mutate(
    OutOfStateShare2020 = BornInOtherState2020 / Population2020 * 100,
    OutOfStateShare2010 = BornInOtherState2010 / Population2010 * 100
  )
```
```{r}
# Now it has geometry + your data, so this will work, lets plot 2010
ggplot(map_data) +
  geom_sf(aes(fill = OutOfStateShare2010), color = "white") +
  scale_fill_viridis_c(option = "D") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 22.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "% of Population Born in Other States in 2010",
    fill = "% Out of State"
  )
```

```{r}
# Now it has geometry + your data, so this will work, lets plot 2020
ggplot(map_data) +
  geom_sf(aes(fill = OutOfStateShare2020), color = "white") +
  scale_fill_viridis_c(option = "D") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 22.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "% of Population Born in Other States in 2020",
    fill = "% Out of State"
  )
```

There is an increase in Merida and its surroundings, but apart from that is not as clear as the Indigenous population. 


## Lets explore the housing pressure

```{r}
housing_pressure_2010 <- biggest_cities_2010 %>%
  mutate(
    POBTOT = as.numeric(P_TOTAL),
    VIVTOT = as.numeric(VIVTOT),
    HousingPressure = POBTOT / VIVTOT
  ) %>%
  select(MUN, NOM_MUN, HousingPressure)

housing_pressure_2020 <- biggest_cities_2020 %>%
  mutate(
    POBTOT = as.numeric(POBTOT),
    VIVTOT = as.numeric(VIVTOT),
    HousingPressure = POBTOT / VIVTOT
  ) %>%
  select(MUN, NOM_MUN, HousingPressure)

# Add a 'Year' column to each and bind them
housing_pressure_2010$Year <- "2010"
housing_pressure_2020$Year <- "2020"

housing_pressure_combined <- bind_rows(housing_pressure_2010, housing_pressure_2020)

## plot both the 2010 and 2020 housing pressure, so they are next to each other
ggplot(housing_pressure_combined, aes(x = reorder(NOM_MUN, -HousingPressure), y = HousingPressure, fill = Year)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Housing Pressure Comparison by Municipality (2010 vs 2020)",
    x = "Municipality",
    y = "People per Inhabited Private Housing Unit",
    fill = "Year"
  ) +
  theme_minimal()
```

It is not easy to distinguish, lets plot it for the cities that have grown the most.
```{r}
most_grown_cities <- population_changes %>%
  filter(PopulationChange > 0) %>%
  arrange(desc(PopulationChange)) %>%
  head(10) %>%
  select(MUN, NOM_MUN)

plot <- housing_pressure_combined %>%
  filter(MUN %in% most_grown_cities$MUN) %>%
  ggplot(aes(x = reorder(NOM_MUN, -HousingPressure), y = HousingPressure, fill = Year)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Housing Pressure in Fastest Growing Municipalities (2010 vs 2020)",
    x = "Municipality",
    y = "People per Inhabited Private Housing Unit",
    fill = "Year"
  ) +
  theme_minimal()
print(plot)
# save to png
ggsave("housing_pressure_fastest_growing.png", plot, width = 10, height = 6, dpi = 300)
```

Lets plot the change in housing units vs population change.
```{r}
population_changes %>%
  left_join(biggest_cities_2020 %>%
              select(MUN, TVIVPARHAB, TVIVPARHAB_2020 = TVIVPARHAB) %>%
              mutate(TVIVPARHAB_2020 = as.numeric(TVIVPARHAB_2020)), 
            by = "MUN") %>%
  left_join(biggest_cities_2010 %>%
              select(MUN, TVIVPARHAB_2010 = TVIVPARHAB) %>%
              mutate(TVIVPARHAB_2010 = as.numeric(TVIVPARHAB_2010)),
            by = "MUN") %>%
  mutate(
    HousingUnitsChange = TVIVPARHAB_2020 - TVIVPARHAB_2010,
    PopulationChange = Population2020 - Population2010
  ) %>%
  ggplot(aes(x = HousingUnitsChange, y = PopulationChange)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Relationship Between Housing Units Change and Population Change",
    x = "Change in Inhabited Private Housing Units",
    y = "Change in Population"
  )
```


It looks like the number of housing units has increased slightly faster than the population.
I checked to see if there's any note in the methodology taken by INEGI, but I did not find anything that suggests the variable does not represent the same anymore.
Some conclusions we can take from this and grounded on my personal knowledge, is that the coast and cities like Merida and Valladolid have become hotspots for people to move into and/ or buy to 
rent out. This could explain the trend, but we would need to do a more in-depth analysis to confirm this.


# Lets test the correlation between the change of out of state migration and the change in indigenous population.
Lets normalize it using z-scores, so we can see the relationship more clearly.
```{r}
# Join the two datasets
correlation_data <- population_changes %>%
  select(MUN, NOM_MUN, BornInOtherStateChange) %>%
  left_join(indigenous_population_changes %>% 
              select(MUN, IndigenousPopulationChange), by = "MUN") %>%
  mutate(
    BornInOtherStateZ = scale(BornInOtherStateChange)[, 1],
    IndigenousChangeZ = scale(IndigenousPopulationChange)[, 1]
  )


# Plot
ggplot(correlation_data, aes(x = BornInOtherStateZ, y = IndigenousChangeZ)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(
    title = "Normalized Relationship Between In-Migration and Indigenous Population Change",
    x = "Z-score of Change in Out-of-State Population",
    y = "Z-score of Change in Indigenous Population"
  ) +
  theme_minimal()
```

It is not clear because the outlier is skewing the results, lets focus on the most grown cities.

```{r}
# Filter for the most grown cities
plot <- ggplot(correlation_data %>%
         filter(MUN %in% most_grown_cities$MUN),
       aes(x = BornInOtherStateZ, y = IndigenousChangeZ)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  geom_text(aes(label = NOM_MUN), vjust = -0.5, size = 3) +
  labs(
    title = "Normalized Relationship Between In-Migration and Indigenous Population Change (Most Grown Cities)",
    x = "Z-score of Change in Out-of-State Population",
    y = "Z-score of Change in Indigenous Population"
  ) +
  theme_minimal()
print(plot)
# save to png
ggsave("normalized_relationship_in_migration_indigenous_change.png", plot, width = 10, height = 6, dpi = 300)
```

Let's test it.
```{r}
cor.test(correlation_data$BornInOtherStateZ, correlation_data$IndigenousChangeZ)
```


I got access to the geojson with localities, so lets first plot the map.

```{r}
localities_map <- st_read("/Users/ecastillo/Documents/coding/analyzing_merida_data/localidades_yucatan_puntos.geojson")
localities_map <- localities_map %>%
  mutate(
    CVE_MUN = as.numeric(cve_mun),  # Ensure CVE_MUN is numeric for joining
    CVE_LOC = as.numeric(cve_loc) # Ensure CVE_LOC is numeric for joining
  )
#plot the map, add black borders and fill with light gray
library(ggplot2)
library(sf)
ggplot(data = localities_map) +
  geom_sf(fill = "lightgray", color = "black") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Localities in Yucatán",
    fill = "Localities"
  )
```

# Now we can join the localities with the population data
```{r}
# Join the localities map with the population data
localities_map <- localities_map %>%
  left_join(data_2020 %>%
              select(MUN, NOM_MUN, LOC, NOM_LOC, POBTOT) %>%
              mutate(MUN = as.numeric(MUN), LOC = as.numeric(LOC)),
            by = c("CVE_MUN" = "MUN", "CVE_LOC" = "LOC"))

```

```{r}
# Plot the localities with population data
ggplot(data = localities_map) +
  geom_sf(aes(fill = POBTOT), color = "black") +
  scale_fill_viridis_c(option = "C") +
  coord_sf(xlim = c(-90.5, -87.5), ylim = c(19.4, 21.5)) +  # Adjust as needed
  theme_minimal() +
  labs(
    title = "Localities in Yucatán with Population Data (2020)",
    fill = "Total Population"
  )
```